Overt attentional capture by reward-related stimuli overcomes inhibitory suppression

### 2019-04-26 analysis/paper/paper.Rmd
references fixed

## Stimuli
The task was based on that used by @Gaspelin2016 and Pearson et al. [-@Pearson2015; -@Pearson2016]. Each trial consisted of a fixation display, a search display, and a feedback display (see Figure \@ref(fig:methodFig)C). All stimuli were presented on a black background. The fixation display consisted of a white cross (0.5$^\circ$ visual angle), presented in the centre of the screen and surrounded by a white circle (diameter 3.0$^\circ$). The search display consisted of four filled shapes (size 1.6$^\circ$ $\times$ 1.6$^\circ$) presented above, below, left and right of screen centre (eccentricity 4.5$^\circ$). One of the shapes was designated as the target, and the other three shapes were designated non-targets. For all participants, the target was randomly determined to be either a circle or a diamond on each trial [^1]. The identity of the non-target shapes varied by search condition. For the *singleton search* condition, the non-target stimuli were all rendered as the alternative target shape (i.e., if the target was a diamond, all of the non-targets were circles; Figure \@ref(fig:methodFig)B). For the *feature search* condition, the non-target stimuli were randomly selected without replacement from a set of two squares and two hexagons (such that one of the non-targets was always a shape singleton), and were never rendered as the alternative target shape (Figure \@ref(fig:methodFig)C). The target and two of the non-targets were always rendered in grey (CIE $x$,$y$ chromaticity coordinates of .327/.400). On the majority of trials, one of the non-target shapes (the *singleton distractor*) was rendered in either orange or blue (CIE $x$,$y$ coordinates .492/.445 and .192/.216, respectively). On the remaining trials, all shapes were rendered in grey such that there was no singleton distractor. The values of blue and orange had similar luminance (~24.5 cd/m^2^), which was higher than that of grey (~8.3 cd/m^2^). The feedback display showed the points earned on the current trial. If the response time (RT) exceeded the soft-timeout threshold (see below), the message "Too slow" appeared below the reward feedback.

>>>>

## Stimuli
The task was based on that used by @Gaspelin2016 and Pearson et al. [-@Pearson2015; -@Pearson2016]. Each trial consisted of a fixation display, a search display, and a feedback display (see Figure \@ref(fig:methodFig)C). All stimuli were presented on a black background. The fixation display consisted of a white cross (0.5$^\circ$ visual angle), presented in the centre of the screen and surrounded by a white circle (diameter 3.0$^\circ$). The search display consisted of four filled shapes (size 1.6$^\circ$ $\times$ 1.6$^\circ$) presented above, below, left and right of screen centre (eccentricity 4.5$^\circ$). One of the shapes was designated as the target, and the other three shapes were designated non-targets. For all participants, the target was randomly determined to be either a circle or a diamond on each trial [^1]. The identity of the non-target shapes varied by search condition. For the *singleton search* condition, the non-target stimuli were all rendered as the alternative target shape (i.e., if the target was a diamond, all of the non-targets were circles; Figure \@ref(fig:methodFig)B). For the *feature search* condition, the non-target stimuli were randomly selected without replacement from a set of two squares and two hexagons (such that one of the non-targets was always a shape singleton), and were never rendered as the alternative target shape (Figure \@ref(fig:methodFig)C). The target and two of the non-targets were always rendered in grey (CIE $x$, $y$ chromaticity coordinates of .327/.400). On the majority of trials, one of the non-target shapes (the *singleton distractor*) was rendered in either orange or blue (CIE $x$, $y$ coordinates .492/.445 and .192/.216, respectively). On the remaining trials, all shapes were rendered in grey such that there was no singleton distractor. The values of blue and orange had similar luminance (~24.5 cd/m^2^), which was higher than that of grey (~8.3 cd/m^2^). The feedback display showed the points earned on the current trial. If the response time (RT) exceeded the soft-timeout threshold (see below), the message "Too slow" appeared below the reward feedback.

### 2019-04-26 analysis/paper/paper.tex
references fixed

Pairing a stimulus with reward is thought to increase its attentional
priority in a way that mimics or otherwise interacts with its physical
salience. For example, eye-tracking studies have demonstrated that VMAC
is largest for the most rapidly generated saccades, and diminishes with
increasing saccade latency (Failing, Nissens, Pearson, Le Pelley, \&
Theeuwes, 2015; Pearson et al., 2016), a pattern similar to that
observed in capture by physical salience (e.g., Zoest, Donk, \&
Theeuwes, 2004). Similarly, electrophysiological studies have shown that
reward affects ERP components thought to index low-level perceptual
activity (Hickey, Chelazzi, \& Theeuwes, 2010). This is consistent with
the idea of \emph{incentive salience} (Berridge \& Robinson, 1998),
wherein a reward-related stimulus becomes more salient and attention
grabbing to the observer through a change in its perceptual
representation. This raises the question, if associating a stimulus with
reward increases its perceptual priority as though it has become more
physically salient, can this priority signal be suppressed by the
inhibitory control process proposed by the signal suppression
hypothesis? That is, can incentive salience be suppressed in the same
way as physical salience? Aside from the theoretical importance of this
question, it may also have important clinical implications. Recent
evidence suggests that the processes underlying VMAC are closely related
to the attentional biases towards drug-related stimuli observed in
addiction (Albertella et al., 2017; Anderson, Faulkner, Rilee, Yantis,
\& Marvel, 2013; for review see Field \& Cox, 2008), which can promote
relapse in recovering addicts (Marhe, Waters, Wetering, \& Franken,
2013; Marissen et al., 2006; Waters, Marhe, \& Franken, 2012). Thus it
is important to know whether cognitive control processes can be used to
overcome the influence of reward on attention when such capture is
counterproductive and contrary to our goals. We addressed this question
by combining Le Pelley et al.'s VMAC procedure with Gaspelin et al.'s
(2017) overt attentional suppression procedure to investigate whether
reward-related attentional biases can be suppressed by inhibitory
control processes under conditions promoting feature search.

@@

The task was based on that used by Gaspelin et al. (2017) and Pearson et
al. (2015, 2016). Each trial consisted of a fixation display, a search
display, and a feedback display (see Figure \ref{fig:methodFig}C). All
stimuli were presented on a black background. The fixation display
consisted of a white cross (0.5\(^\circ\) visual angle), presented in
the centre of the screen and surrounded by a white circle (diameter
3.0\(^\circ\)). The search display consisted of four filled shapes (size
1.6\(^\circ\) \(\times\) 1.6\(^\circ\)) presented above, below, left and
right of screen centre (eccentricity 4.5\(^\circ\)). One of the shapes
was designated as the target, and the other three shapes were designated
non-targets. For all participants, the target was randomly determined to
be either a circle or a diamond on each trial.\footnote{This aspect of
  our design differs from that of Gaspelin et al. (2017). In their
  study, for half of participants in the feature search condition the
  target was always a circle, and for other participants the target was
  always a diamond. By contrast, for the singleton search condition, for
  all participants the target was a circle among diamonds on some
  trials, and a diamond among circles on others. That is, target
  identity (circle/diamond) was manipulated between-subjects for the
  feature search condition, and within-subjects for the singleton search
  condition. In the current study, we equated the two conditions by
  varying target identity within-subjects for all participants. This
  ensured that any difference between the two search conditions was
  driven by a difference in the specificity of participants' target
  template and not by a difference in participants' history of selecting
  the target shape (e.g., Maljkovic \& Nakayama, 1994).} The identity of
the non-target shapes varied by search condition. For the
\emph{singleton search} condition, the non-target stimuli were all
rendered as the alternative target shape (i.e., if the target was a
diamond, all of the non-targets were circles; Figure
\ref{fig:methodFig}B). For the \emph{feature search} condition, the
non-target stimuli were randomly selected without replacement from a set
of two squares and two hexagons (such that one of the non-targets was
always a shape singleton), and were never rendered as the alternative
target shape (Figure \ref{fig:methodFig}C). The target and two of the
non-targets were always rendered in grey (CIE \(x\),\(y\) chromaticity
coordinates of .327/.400). On the majority of trials, one of the
non-target shapes (the \emph{singleton distractor}) was rendered in
either orange or blue (CIE \(x\),\(y\) coordinates .492/.445 and
.192/.216, respectively). On the remaining trials, all shapes were
rendered in grey such that there was no singleton distractor. The values
of blue and orange had similar luminance (\textasciitilde{}24.5
cd/m\textsuperscript{2}), which was higher than that of grey
(\textasciitilde{}8.3 cd/m\textsuperscript{2}). The feedback display
showed the points earned on the current trial. If the response time (RT)
exceeded the soft-timeout threshold (see below), the message
\enquote{Too slow} appeared below the reward feedback.

@@

An inconsistency between the results of the current study and those of
Gaspelin et al. (2017) comes from the time-course of the attentional
capture and suppression effects that were observed. According to the
signal suppression hypothesis, suppression is \emph{proactive}---in that
it operates prior to the initial allocation of attention---and so
suppression should be observed even for the most rapidly generated
saccades. Consistent with this idea, Gaspelin et al. (2017) observed an
overt attentional suppression effect to physically salient singleton
distractors amongst the fastest quartile of saccade latencies. By
contrast, in the current experiment we observed overt attentional
capture in the fastest quartile of saccade latencies, with the
suppression effect emerging on trials with slower saccades.
Methodological differences between Gaspelin et al.\enquote{s (2017)
experiment and the current study may be responsible for this
discrepancy. For example, the availability of reward on low-reward
trials may have influenced capture, either by somewhat increasing the
attentional priority of the stimulus feature associated with low reward,
or by otherwise influencing participants} motivation to respond quickly.
Alternatively, the discrepancy in findings may reflect our use of two
potential targets (circle and diamond) in the feature search condition,
as opposed to the same target on all trials in Gaseplin et al.'s
procedure. Having two possible targets could reduce the influence of
trial-by-trial upweighting of target features (Maljkovic \& Nakayama,
1994), but may also result in less effective feature search (but see
Grubert \& Eimer, 2015; e.g., Houtkamp \& Roelfsema, 2009). Another
possible explanation comes from the use of an intermixed trial
structure: as participants randomly experienced trials containing either
the high- or low-reward colour throughout the course of the task, it is
possible that the proactive down-weighting of the salient-but-irrelevant
features could not build up enough to prevent capture for the most
rapidly initiated saccades (Gaspelin \& Luck, 2019). How these
methodological differences might influence the time-course of overt
attentional capture and suppression remains a question for future
research. Nonetheless, the finding does raise questions about whether
the attentional suppression effect that we have observed is truly the
result of proactive attentional suppression of salient stimuli, as
suggested by the signal suppression hypothesis, or potentially the
result of rapid covert disengagement from the salient stimuli (see
Theeuwes, 2010; Zoest et al., 2004 for similar arguments).

@@

@@

\hypertarget{ref-VanZoest2004}{}
Zoest, W. van, Donk, M., \& Theeuwes, J. (2004). The role of
stimulus-driven and goal-driven control in saccadic visual selection.
\emph{Journal of Experimental Psychology: Human Perception and
Performance}, \emph{30}, 746--759.
\url{https://doi.org/10.1037/0096-1523.30.4.749}

>>>>

Pairing a stimulus with reward is thought to increase its attentional
priority in a way that mimics or otherwise interacts with its physical
salience. For example, eye-tracking studies have demonstrated that VMAC
is largest for the most rapidly generated saccades, and diminishes with
increasing saccade latency (Failing, Nissens, Pearson, Le Pelley, \&
Theeuwes, 2015; Pearson et al., 2016), a pattern similar to that
observed in capture by physical salience (e.g., van Zoest, Donk, \&
Theeuwes, 2004). Similarly, electrophysiological studies have shown that
reward affects ERP components thought to index low-level perceptual
activity (Hickey, Chelazzi, \& Theeuwes, 2010). This is consistent with
the idea of \emph{incentive salience} (Berridge \& Robinson, 1998),
wherein a reward-related stimulus becomes more salient and attention
grabbing to the observer through a change in its perceptual
representation. This raises the question, if associating a stimulus with
reward increases its perceptual priority as though it has become more
physically salient, can this priority signal be suppressed by the
inhibitory control process proposed by the signal suppression
hypothesis? That is, can incentive salience be suppressed in the same
way as physical salience? Aside from the theoretical importance of this
question, it may also have important clinical implications. Recent
evidence suggests that the processes underlying VMAC are closely related
to the attentional biases towards drug-related stimuli observed in
addiction (Albertella et al., 2017; Anderson, Faulkner, Rilee, Yantis,
\& Marvel, 2013; for review see Field \& Cox, 2008), which can promote
relapse in recovering addicts (Marhe, Waters, Wetering, \& Franken,
2013; Marissen et al., 2006; Waters, Marhe, \& Franken, 2012). Thus it
is important to know whether cognitive control processes can be used to
overcome the influence of reward on attention when such capture is
counterproductive and contrary to our goals. We addressed this question
by combining Le Pelley et al.'s VMAC procedure with Gaspelin et al.'s
(2017) overt attentional suppression procedure to investigate whether
reward-related attentional biases can be suppressed by inhibitory
control processes under conditions promoting feature search.

@@

The task was based on that used by Gaspelin et al. (2017) and Pearson et
al. (2015, 2016). Each trial consisted of a fixation display, a search
display, and a feedback display (see Figure \ref{fig:methodFig}C). All
stimuli were presented on a black background. The fixation display
consisted of a white cross (0.5\(^\circ\) visual angle), presented in
the centre of the screen and surrounded by a white circle (diameter
3.0\(^\circ\)). The search display consisted of four filled shapes (size
1.6\(^\circ\) \(\times\) 1.6\(^\circ\)) presented above, below, left and
right of screen centre (eccentricity 4.5\(^\circ\)). One of the shapes
was designated as the target, and the other three shapes were designated
non-targets. For all participants, the target was randomly determined to
be either a circle or a diamond on each trial.\footnote{This aspect of
  our design differs from that of Gaspelin et al. (2017). In their
  study, for half of participants in the feature search condition the
  target was always a circle, and for other participants the target was
  always a diamond. By contrast, for the singleton search condition, for
  all participants the target was a circle among diamonds on some
  trials, and a diamond among circles on others. That is, target
  identity (circle/diamond) was manipulated between-subjects for the
  feature search condition, and within-subjects for the singleton search
  condition. In the current study, we equated the two conditions by
  varying target identity within-subjects for all participants. This
  ensured that any difference between the two search conditions was
  driven by a difference in the specificity of participants' target
  template and not by a difference in participants' history of selecting
  the target shape (e.g., Maljkovic \& Nakayama, 1994).} The identity of
the non-target shapes varied by search condition. For the
\emph{singleton search} condition, the non-target stimuli were all
rendered as the alternative target shape (i.e., if the target was a
diamond, all of the non-targets were circles; Figure
\ref{fig:methodFig}B). For the \emph{feature search} condition, the
non-target stimuli were randomly selected without replacement from a set
of two squares and two hexagons (such that one of the non-targets was
always a shape singleton), and were never rendered as the alternative
target shape (Figure \ref{fig:methodFig}C). The target and two of the
non-targets were always rendered in grey (CIE \(x\), \(y\) chromaticity
coordinates of .327/.400). On the majority of trials, one of the
non-target shapes (the \emph{singleton distractor}) was rendered in
either orange or blue (CIE \(x\), \(y\) coordinates .492/.445 and
.192/.216, respectively). On the remaining trials, all shapes were
rendered in grey such that there was no singleton distractor. The values
of blue and orange had similar luminance (\textasciitilde{}24.5
cd/m\textsuperscript{2}), which was higher than that of grey
(\textasciitilde{}8.3 cd/m\textsuperscript{2}). The feedback display
showed the points earned on the current trial. If the response time (RT)
exceeded the soft-timeout threshold (see below), the message
\enquote{Too slow} appeared below the reward feedback.

@@

An inconsistency between the results of the current study and those of
Gaspelin et al. (2017) comes from the time-course of the attentional
capture and suppression effects that were observed. According to the
signal suppression hypothesis, suppression is \emph{proactive}---in that
it operates prior to the initial allocation of attention---and so
suppression should be observed even for the most rapidly generated
saccades. Consistent with this idea, Gaspelin et al. (2017) observed an
overt attentional suppression effect to physically salient singleton
distractors amongst the fastest quartile of saccade latencies. By
contrast, in the current experiment we observed overt attentional
capture in the fastest quartile of saccade latencies, with the
suppression effect emerging on trials with slower saccades.
Methodological differences between Gaspelin et al.\enquote{s (2017)
experiment and the current study may be responsible for this
discrepancy. For example, the availability of reward on low-reward
trials may have influenced capture, either by somewhat increasing the
attentional priority of the stimulus feature associated with low reward,
or by otherwise influencing participants} motivation to respond quickly.
Alternatively, the discrepancy in findings may reflect our use of two
potential targets (circle and diamond) in the feature search condition,
as opposed to the same target on all trials in Gaseplin et al.'s
procedure. Having two possible targets could reduce the influence of
trial-by-trial upweighting of target features (Maljkovic \& Nakayama,
1994), but may also result in less effective feature search (but see
Grubert \& Eimer, 2015; e.g., Houtkamp \& Roelfsema, 2009). Another
possible explanation comes from the use of an intermixed trial
structure: as participants randomly experienced trials containing either
the high- or low-reward colour throughout the course of the task, it is
possible that the proactive down-weighting of the salient-but-irrelevant
features could not build up enough to prevent capture for the most
rapidly initiated saccades (Gaspelin \& Luck, 2019). How these
methodological differences might influence the time-course of overt
attentional capture and suppression remains a question for future
research. Nonetheless, the finding does raise questions about whether
the attentional suppression effect that we have observed is truly the
result of proactive attentional suppression of salient stimuli, as
suggested by the signal suppression hypothesis, or potentially the
result of rapid covert disengagement from the salient stimuli (see
Theeuwes, 2010; van Zoest et al., 2004 for similar arguments).

@@

\hypertarget{ref-VanZoest2004}{}
van Zoest, W., Donk, M., \& Theeuwes, J. (2004). The role of
stimulus-driven and goal-driven control in saccadic visual selection.
\emph{Journal of Experimental Psychology: Human Perception and
Performance}, \emph{30}, 746--759.
\url{https://doi.org/10.1037/0096-1523.30.4.749}

@@
